Graphical Tests for Power Comparison of Competing Designs

### 2012-03-29 setup.tex
table re-worked

For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for all three of our response values, using the {\t R} package {\tt lme4} \cite{bates:2011}. 
For power predictions (cf. table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates of {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and as two-way interaction effects with design to assess their effect on the power of designs. To adjust for individuals' ability, a random intercept was included in the model.

@@

Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes. The reference line has surprisingly little influence, but it helps more for  polar charts than for barcharts,
An increase in sample size has positive impact on power (see figure \ref{fig:treatment} for an overview). Polar charts need much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The offset changes are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. Polar charts benefit more from this inversion than barcharts. This is an unexpected finding, but is consistent throughout different lineups in the data. Figure \ref{fig:power} summarizes the results from the model: power predictions are shown by sample size ($x$ axis). The thick lines show average power by design for different shifts in wind direction. The thin lines represent power for individuals. What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the 'mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.

@@

 Confidence levels are measured on a five point scale --- they are subjective assessments by the participant `how certain are you'. We model the value using a normal distribution - a discrete uniform distribution  might be more appropriate,
While there is a difference between the designs - participants reported higher confidence in dealing with the barcharts -- this is only  a trend (i.e. not significant at 5\% but below 10\%). The only significant effect on confidence level is visible for the use of reference lines in polar coordinates: participants report an increase in confidence when using the reference line (however, as we have seen before, this is not matched by an increase in actual correctness of the results).

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
%\rowcolors{2}{white}{lightgray}
\begin{tabular}{rrrrrrl}
  \hline
& & Estimate & Error & $z$-value & $p$-value &\\ 
  \hline
\bf design & euclid & -0.08 & 0.39 & -0.21 & 0.84 &\\ 
&polar & -1.98 & 0.32 & -6.13 & 0.00 & ***\\ [3pt]
\multicolumn{2}{l}{\bf main effects} &&&&&\\
& reference line  & -0.14 & 0.26 & -0.53 & 0.59 & \\ [1pt]
&  sample size & 0.27 & 0.04 & 6.31 & 0.00 & ***\\[1pt]
 &offset:  90 degrees& -0.43 & 0.37 & -1.18 & 0.24 &\\ 
  & 180 degrees& -0.89 & 0.35 & -2.51 & 0.01 & **\\ 
  & 270 degrees& 0.21 & 0.38 & 0.55 & 0.58 &\\ [3pt]
\multicolumn{2}{l}{\bf interactions} &&&&&\\
&  polar:line & 0.51 & 0.35 & 1.44 & 0.15 &\\ [1pt]
&    polar:sample size & -0.23 & 0.05 & -5.02 & 0.00 & ***\\[1pt]
&    polar:offset 90 & 0.64 & 0.49 & 1.30 & 0.20 \\ 
&  polar:offset 180 & 0.91 & 0.47 & 1.92 & 0.05 & .\\ 
&    polar:offset 270 & -0.73 & 0.54 & -1.35 & 0.18 &\\
   \hline
\end{tabular}
}
\end{center}
\caption{\label{tbl:correct} Output of a generalized linear mixed effects model for power of lineups (i.e. probability of identifying the data plot) for comparing designs. Included are two-way interactions with sample size and shifts in wind direction (offset). Results are based on  976 lineup evaluations by 115 participants. }
\end{table}

>>>>

For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for all three of our response values, using the {\t R} package {\tt lme4} \cite{bates:2011}. 
For power predictions (cf. table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and in two-way interaction effects with design to assess their effect on the power of designs. To adjust for individuals' ability, a random intercept was included in the model.

@@

Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes. The reference line has surprisingly little influence, but it helps more for  polar charts than for barcharts,
An increase in sample size has positive impact on power (see figure \ref{fig:treatment} for an overview). Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. The Power of Barcharts suffers significantly  from this inversion whereas power of polar charts is unaffected. This is an unexpected finding, but is consistent throughout different lineups in the data. Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis). The thick lines show average power by design for different shifts in wind direction. The thin lines represent power for individuals. What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the 'mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.

@@

 Confidence levels are measured on a five point scale --- they are subjective assessments by the participant `how certain are you'. We model confidence using the same model structure as before, i.e. using sample size and offset as covariates and including up to two-way interactions,
While there is a difference between the designs - participants reported higher confidence in dealing with the barcharts -- this is only  a trend (i.e. not significant at 5\% but below 10\%). The only significant effect on reported confidence level is the use of reference lines in polar coordinates: participants report an increase in confidence (0.31 $\pm$ 0.13, $p$ value = 0.01) when using the reference line. This, however, does not translate to a significant increase in accuracy of results, as we have seen before.

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
%\rowcolors{2}{white}{lightgray}
\begin{tabular}{rrrrrrl}
  \hline
& & Estimate & Error & $z$-value & $p$-value &\\ 
  \hline
\bf design & euclid & -0.08 & 0.39 & -0.21 & 0.84 &\\ 
&polar & -1.98 & 0.32 & -6.13 & 0.00 & ***\\ [3pt]
\multicolumn{2}{l}{\bf main effects} &&&&&\\
& reference line  & -0.14 & 0.26 & -0.53 & 0.59 & \\ [1pt]
&  sample size & 0.27 & 0.04 & 6.31 & 0.00 & ***\\[1pt]
 &offset:  90 degrees& -0.43 & 0.37 & -1.18 & 0.24 &\\ 
  & 180 degrees& -0.89 & 0.35 & -2.51 & 0.01 & **\\ 
  & 270 degrees& 0.21 & 0.38 & 0.55 & 0.58 &\\ [3pt]
\multicolumn{2}{l}{\bf interactions} &&&&&\\
&  polar:line & 0.51 & 0.35 & 1.44 & 0.15 &\\ [1pt]
&    polar:sample size & -0.23 & 0.05 & -5.02 & 0.00 & ***\\[1pt]
&    polar:offset 90 & 0.64 & 0.49 & 1.30 & 0.20 \\ 
&  polar:offset 180 & 0.91 & 0.47 & 1.92 & 0.05 & .\\ 
&    polar:offset 270 & -0.73 & 0.54 & -1.35 & 0.18 &\\
   \hline
\\[-5pt]
   \multicolumn{5}{l}{Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1}
\end{tabular}
}
\end{center}
\caption{\label{tbl:correct} Output of a generalized linear mixed effects model for power of lineups (i.e. probability of identifying the data plot) for comparing designs. Included are two-way interactions with sample size and shifts in wind direction (offset). Results are based on  976 lineup evaluations by 115 participants. }
\end{table}
